---
title: "PS270 Final Project"
author: "Jane C. Olson"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in looking at data analyzing what parties American voters tend to say they prefer in presidential elections as they come closer to the election. I am curious as to whether or not voters tend to favor a certain party in the beginning of the election cycle versus the period nearest to the election. I hope to see this through American presidential pre-election polls.

## Tentative Proposal

Are American voters more attracted to a certain party as the presidential election comes closer? In this study, I plan to examine pre-election polls regarding American presidential elections. I hypothesize that as a presidential election draws nearer, a greater amount of polls will favor the candidate of one of the two main parties. When there is more time between polls and the presidential election, I suspect that polls will be more spread out in their predictions of which candidate will be elected. My sample is comprised of pre-election polls from American presidential elections between 1942 and 2017. My unit of analysis is the results of the pre-election polls from one of the elections between 1942 and 2017 (e.g., the 2016 presidential election). The explanatory variable of interest is the amount of days away from the election the poll is taken. The variable is coded in accordance with the amount of days before the election day. For instance, if the variable were coded =1462, it would mean that the poll was taken 1,462 days before election day. My outcome variable is the candidate who is predicted to win the election in the pre-election polls, specifically which party they run under. The variable is coded =1 for the Democratic candidate and =2 for the Republican candidate. This variable is measured from Christopher Wlezien's data set on national polls assessing citizens' vote intentions in 45 countries between 1942 and 2017. If I observe a greater amount of polls predicting one candidate to win over the other (e.g., more polls predict the Democratic candidate to win than the Republican candidate) closer to the election date, this would provide support for my hypothesis. If, on the other hand, I observe more even amounts of polls predicting either the Democratic or Republican candidate to win, this would provide evidence against my hypothesis. When I run my regression of the predictions of pre-election polls on the days before election day variable, a positive, significant coefficient closer to 1 or 2 would indicate support for my hypothesis.

## Data Visualization

```{r}
library(tidyverse)
us_pres_polls_2016 <- read_tsv("final-data/Dataset-20180111.tab") |>
  filter(country == "United States",
         election == "Presidential",
         electionyr == "2016")

pres_2016_plot <- ggplot(us_pres_polls_2016, mapping = aes(x = polldate,
                                                                y = poll_)) +
  geom_point(mapping = aes(color = partyid)) +
  facet_wrap(~ partyid) +
  labs(
        x = "Date of Poll",
        y = "Predicted Percentage of Votes",
        title = "Expected Percentage of Votes for Each 2016 Presidential Candidate",
        subtitle = "Based on Dates of Polls",
        caption = "Will Jennings & Christopher Wlezien"
      ) +
  theme(panel.border = element_blank(), panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())
pres_2016_plot
```

## Analysis

The graphs below show that as the time between the 2012 election and polling decreased, the amount of votes predicted for both the Democratic and Republican candidates increased. This indicates that support did not shift toward one individual candidate and instead, support in the polls increased because of other factors. One of these factors could be that the amount of polls taken increased as the amount of days until the election decreased, so there were more opportunities for those included in the polls to say they planned to vote for either the Democratic or Republican candidate.

```{r}
us_pres_polls_2012 <- read_tsv("final-data/Dataset-20180111.tab") |>
  filter(country == "United States",
         election == "Presidential",
         electionyr == "2012")

pres_2012_plot <- ggplot(us_pres_polls_2012, mapping = aes(x = polldate,
                                                                y = poll_)) +
  geom_point(mapping = aes(color = partyid)) +
  facet_wrap(~ partyid) +
  labs(
        x = "Date of Poll",
        y = "Predicted Percentage of Votes",
        title = "Expected Percentage of Votes for Each 2012 Presidential Candidate",
        subtitle = "Based on Dates of Polls",
        caption = "Will Jennings & Christopher Wlezien"
      ) +
  geom_smooth(color = "red") +
  theme(panel.border = element_blank(), panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())
pres_2012_plot
```





